From Reversible Programs to \\ Univalent Universes and Back

### 2017-05-12 upi.tex
more on univalent fibrations

As a simple example of a non-univalent fibration consider $P : \mathbb{2}
\rightarrow \mathbb{U}$ defined as $P = \lambda \_. \mathbb{1}$ illustrated below:

\bigskip
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $\mathbb{2}$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$\fc$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$\tc$};
  \draw (0,-2.7) ellipse (0.5cm and 1cm);
  \node[right] at (0.5,-2.7) {Fiber $P(\fc)$ = Fiber $P(\tc)$ = $\mathbb{1}$};
  \draw[fill] (0,-2.7) circle [radius=0.025];
  \node[below] at (0,-2.7) {$\star$};
\end{tikzpicture}

\medskip\noindent There is one equivalence
$P(\fc) = \mathbb{1} \simeq \mathbb{1} = P(\tc)$ but this certainly does not
induce a path $\fc \equiv \tc$.

>>>>

As two simple examples of non-univalent fibrations consider
$P : \mathbb{1} \rightarrow \mathbb{U}$ defined as $P = \lambda\_.\mathbb{2}$
and $Q : \mathbb{2} \rightarrow \mathbb{U}$ defined as
$Q = \lambda \_. \mathbb{1}$ illustrated below:

\begin{tabular}{c@{\qquad\qquad}c}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $\mathbb{1}$};
  \draw[fill] (0,-5) circle [radius=0.025];
  \node[below] at (0,-5) {$\star$};
  \draw (0,-2.7) ellipse (0.5cm and 1cm);
  \node[right] at (0.5,-2.7) {Fiber $P(\star) = \mathbb{2}$};
  \draw[fill] (-0.2,-2.7) circle [radius=0.025];
  \draw[fill] (0.2,-2.7) circle [radius=0.025];
\end{tikzpicture}
&
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $\mathbb{2}$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$\fc$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$\tc$};
  \draw (0,-2.7) ellipse (0.5cm and 1cm);
  \node[right] at (0.5,-2.7) {Fiber $Q(\fc)$ = Fiber $Q(\tc)$ = $\mathbb{1}$};
  \draw[fill] (0,-2.7) circle [radius=0.025];
  \node[below] at (0,-2.7) {$\star$};
\end{tikzpicture}
\end{tabular}

\medskip\noindent The identity function and boolean negation induce two distinct
equivalences in the space $P(\star) = \mathbb{2} \simeq \mathbb{2} =
P(\star)$. As there is only one path $\star \equiv star$ in the base space, the
fibration $P$ is not univalent. Similarly, there is one equivalence
$Q(\fc) = \mathbb{1} \simeq \mathbb{1} = Q(\tc)$ but this certainly does not
induce a path $\fc \equiv \tc$. As the examples illustrate there is a delicate
balance required for the fibration to be univalent; otherwise some distinct
equivalences would be identified (like for $P$) or distinct points in the base
space would be identified (like for $Q$).

An example of a univalent fibration is $P : \mathbb{2} \rightarrow \mathbb{U}$
defined as $P(\fc) = \mathbb{0}$ and $P(\tc) = \mathbb{1}$ illustrated below:

\bigskip
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $\mathbb{2}$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$\fc$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$\tc$};
  \draw (-1.3,-2) ellipse (0.5cm and 2cm);
  \node[left] at (-1.8,-2) {Fiber $P(\fc) = \mathbb{0}$};
  \draw (1.3,-2) ellipse (0.5cm and 2cm);
  \node[right] at (1.8,-2) {Fiber $P(\tc) = \mathbb{1}$};
  \draw[fill] (1.8,-2) circle [radius=0.025];
  \draw[below,cyan,dashed,thick] (-1,-5) -- (1,-5);
  \node[below,cyan,dashed,thick] at (0,-5) {$\equiv$};
  \draw[->,red,dashed,ultra thick] (-0.8,-1.7) to [out=45, in=135] (0.8,-1.7);
  \draw[->,red,dashed,ultra thick] (0.8,-2.3) to [out=-135, in=-45] (-0.8,-2.3);
  \node[red,ultra thick] at (0,-2) {$\simeq$};
\end{tikzpicture}

\medskip\noindent In this case, the space of equivalences
$P(\fc) = \mathbb{0} \simeq \mathbb{1} = P(\tc)$ is empty and so is the space of
paths $\fc \equiv \tc$ and hence these two spaces are vacuously equivalent.
